---
title: "Report"
description: Visualizing diversity in research outputs at the University of Basel
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(ggtext)
library(patchwork)
library(readxl)

pub_data <- read_csv("1_data/unibas_diversity.csv")
prof_data <- read_xlsx("1_data/akademisches_unibas.xlsx", sheet = "female_perc")
```

## Executive Summary

Monitoring of diversity is an important first step towards reducing inequalities in access to scientific innovation and research. We contribute to monitoring of author diversity in scientific production at the University of Basel by conducting an analysis of gender distribution in scientific publications. We analysed the diversity of the University of Basel's research output by categorizing authors' gender based on authors' first names and visualizing temporal trends between 2000 and 2020 in gender distribution for all 7 faculties of the University of Basel. Our results for the University of Basel match similar reports of gender publication gaps in the sciences. Crucially, we find large differences across faculties and author positions (first vs. last) suggesting that targeted measures may be needed for different organizational units and career stages. We discuss implications for future efforts to monitoring diversity in scientific production and associated interventions at the University of Basel and beyond.
 
## Background

#### Why does diversity matter?

Achieving gender equality (SDG 5) and reducing inequalities (SDG10) are central parts of the [Sustainable Development Goals (SDGs)](https://sdgs.un.org/goals). Universities can play a key role in advancing these goals by empowering women and disadvantaged minorities to allow them to make contributions to scientific innovation and research. The University of Basel is strongly committed to monitoring and increasing diversity of its staff and student population in a number of efforts spearheaded by its [Diversity Office](https://www.unibas.ch/en/University/Administration-Services/Vice-President-s-Office-for-People-and-Culture/Diversity-and-Inclusion/Diversity.html). 

#### Why does monitoring of research outputs matter?

Monitoring diversity is a first step towards identifying inequalities and potential solutions to reduce them. As a consequence, there have been continued efforts on the part of the [Diversity Office](https://www.unibas.ch/en/University/Administration-Services/Vice-President-s-Office-for-People-and-Culture/Diversity-and-Inclusion/Diversity.html) of the University of Basel to monitor diversity, for example, in the form of an [Equal opportunities monitoring](https://www.unibas.ch/en/University/Administration-Services/Vice-President-s-Office-for-People-and-Culture/Diversity-and-Inclusion/Diversity/Diversity-Management/Monitoring.html). However, the current monitoring does not fully capture all aspects of equal opportunities to scientific production. The monitoring of publication outputs can be helpful as an additional indicator because it has the potential to reflect access to resources (e.g., time, internal funding) that go beyond employment and are otherwise not easily quantifiable. 

## Goals of this Report
- visualize the gender diversity in research outputs at the University Basel
- provide ideas for future monitoring of diversity at the University of Basel, including future methodological approaches or data collection efforts
- discuss whether the monitoring of research outputs can help guide future interventions aimed at increasing diversity at the University of Basel

## Methods 

#### Research Output of the University of Basel

We use data from a database listing the University of Basel's research output. We focus on data between 2000 and 2020 because availability for earlier years is very limited and this time range allows some comparability to previous efforts (e.g., Odic & Wojcik, 2019). We focus our analyses on University of Basel's 7 faculties and neglect smaller associated institutes because there is large number of the latter with relatively few publications that make it difficult to consider trends over time.

We analyze a total 67'269 publications including several different categories: 43'997 journal articles, 11'959 book chapters, 3'454 books (edited and authored), 2'014 proceedings articles, 1'671 theses, and 4'174 other items including news articles and discussion papers.    

#### Employment Information of the University of Basel

We use data from the [Equal opportunities monitoring](https://www.unibas.ch/en/University/Administration-Services/Vice-President-s-Office-for-People-and-Culture/Diversity-and-Inclusion/Diversity/Diversity-Management/Monitoring.html) concerning the percentage of females in different positions (i.e., professorships, academic staff) available for the period 2013-2019. 

#### Diversity Labeling from Author First Names

We use information concerning diversity derived from the authors' first names. We recruited existing publicly available services (e.g., [genderize.io](https://genderize.io/)) to determine each author's individual characteristics, such as gender. Estimates are provided based on the services' databases that include hundreds of thousands of confirmed mappings between first names and individuals' characteristics. Additional details are provided on the [Data](data.html) page. 

Our report focuses on gender in line with the majority of past monitoring of diversity at the University of Basel and other similar efforts. Ideally, however, diversity should be conceptualized through several dimensions, that could include cultural and socio-economic background, political views, age, nationality, or religion, to name but a few. In our efforts, we strove to include different dimensions beyond gender, namely, nationality and age. Our simple validation efforts suggested, however, that such categorizations are unreliable so we refrain from reporting results concerning these dimensions below. We discuss limitations of our approach and alternative methods in the *Limitations and Outlook* section. 

#### Disclaimer

Our analyses used first names to estimate the gender of authors using extant technology. This classification system assigns names to either "female" or "male", thus, providing a traditional, binary classification of gender. The contributors would like to state that this choice is of technical and pragmatic nature and should not be understood as a rejection of more inclusive, non-binary gender definitions. 

```{r, echo=FALSE, eval=TRUE}
# OVER TIME GENDER BY FACULTY ----------------------------------------------------------------

# data
gender_inst_time <- pub_data %>% 
  select(year, gender_proportion, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:gender_proportion,names_to = "organisation_name", values_to = "applies") %>% 
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Theology",
                                       organisation_name == "Org-Faculty of Science" ~ "Natural Sciences",
                                       organisation_name == "Org-Faculty of Law" ~ "Law",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Humanities &\nSocial Sciences",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Medicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Business &\nEconomics",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Psychology")) %>% 
  filter(!is.na(organisation_name)) %>% # ignoring non-faculty organizations
  group_by(organisation_name, year) %>% 
  summarise(FEMALE = 100*mean(gender_proportion, na.rm = TRUE),
            MALE = 100 - FEMALE) %>%
  pivot_longer(FEMALE:MALE, names_to = "gender", values_to = "publication")

gender_inst_time$gender = factor(gender_inst_time$gender,levels = c("MALE","FEMALE"))


# Plot

charcoal <-  "#31424B"

label_num <- data.frame(organisation_name= unique(gender_inst_time$organisation_name), label= c("50%", rep("", 6)))
label_f <- data.frame(organisation_name= unique(gender_inst_time$organisation_name), label= c("FEMALE", rep("", 6)))
label_m <- data.frame(organisation_name= unique(gender_inst_time$organisation_name), label= c("MALE", rep("", 6)))

pA1 <- ggplot(gender_inst_time, aes(x= year, y=publication, fill=gender)) +
  geom_bar(width = 1,position = "stack", stat = "identity") +
  scale_fill_manual(values = c("FEMALE" = "#4ADABF", "MALE" = "#4E4BD8"))+
  geom_hline(yintercept = 50, linetype = "dashed", color = "white") +
  theme_void() +
  geom_label(data = label_num, aes(label=label), fill = NA,
             x = 2001, y = 51, hjust= 0, vjust = 0,  label.size = NA,
             family = "Oswald Bold", size = 7, color = "white" ,
             inherit.aes = FALSE) +
  geom_label(data = label_f, aes(label=label), fill = NA,
             y = 5, x = 2019.75, hjust= 1,  label.size = NA,
             family = "Oswald Bold", size = 9, color = "#4E4BD8" ,
             inherit.aes = FALSE) +
  geom_label(data = label_m, aes(label=label), fill = NA,
             y = 95, x = 2019.75, hjust= 1,  label.size = NA,
             family = "Oswald Bold", size = 9, color ="#4ADABF" ,
             inherit.aes = FALSE) +
  scale_x_continuous(breaks = c(2000, 2020)) +
  scale_y_continuous(expand = c(0.01,0), breaks = c(0, 100)) +
  facet_wrap(.~ toupper(organisation_name), nrow = 1) +
  labs(title = "PERCENTAGE OF FEMALE AUTHORS (2000 - 2020)", 
       x = "YEAR", y = "PERCENTAGE",
       caption = "Data from the University of Basel.") +
  theme(legend.position = "none",
        axis.text.x = element_text(family = "Oswald", size = 18, colour = "grey60"),
        axis.text.y = element_text(family = "Oswald", size = 18, colour = "grey60"),
        panel.spacing = unit(1, "lines"),
        strip.text = element_text(family = "Oswald Bold", size = 18, colour = charcoal),
        axis.title.x = element_text(family = "Oswald Bold", size = 18, colour = "grey60"),
        axis.title.y = element_text(family = "Oswald Bold", size = 18, colour = "grey60", angle = 90),
        plot.title = element_text(face = "bold", family = "Oswald", hjust = 0, margin = margin(t = 40,b = 25), size = 24, color = charcoal),
        plot.subtitle = element_text(family = "Oswald", hjust = 0.008, margin = margin(b = 10), size = 17, color = charcoal),
        plot.caption =element_text(family = "Oswald ExtraLight", size = 20, hjust = 1, margin = margin(t = 20, b = 3, r = 3)))



# OVER TIME FEMALE 1ST and LAST AUTHOR BY FACULTY ----------------------------------------------------------------

# data


author_first <- pub_data %>% 
  filter(n_authors > 1) %>% 
  select(year, gender_first, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:gender_first,names_to = "organisation_name", values_to = "applies") %>% 
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Theology",
                                       organisation_name == "Org-Faculty of Science" ~ "Natural Sciences",
                                       organisation_name == "Org-Faculty of Law" ~ "Law",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Humanities &\nSocial Sciences",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Medicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Business &\nEconomics",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Psychology")) %>% 
  filter(!is.na(organisation_name)) %>% # ignoring non-faculty organizations
  group_by(organisation_name, year) %>% 
  summarise(mean_f = 100*mean(gender_first, na.rm = TRUE))  %>% 
  mutate(author_type = "FIRST AUTHOR") 



author_last <- pub_data %>% 
  filter(n_authors > 1) %>% 
  select(year, gender_last, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:gender_last,names_to = "organisation_name", values_to = "applies") %>% 
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Theology",
                                       organisation_name == "Org-Faculty of Science" ~ "Natural Sciences",
                                       organisation_name == "Org-Faculty of Law" ~ "Law",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Humanities &\nSocial Sciences",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Medicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Business &\nEconomics",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Psychology")) %>% 
  filter(!is.na(organisation_name)) %>% # ignoring non-faculty organizations
  group_by(organisation_name, year) %>% 
  summarise(mean_f = 100*mean(gender_last, na.rm = TRUE))  %>% 
  mutate(author_type = "LAST AUTHOR") 

author_dat <- bind_rows(author_first, author_last) 


label_num <- data.frame(organisation_name= unique(author_dat$organisation_name), label= c("50%", rep("", 6)))
label_f <- data.frame(organisation_name= unique(author_dat$organisation_name), label= c("FIRST AUTHOR", rep("", 6)))
label_l <- data.frame(organisation_name= unique(author_dat$organisation_name), label= c("LAST AUTHOR", rep("", 6)))


# plot

charcoal <-  "#31424B"
col_pal <- c("#4ADABF", "white")

pA2 <- author_dat %>% 
  ggplot(aes(x = year, y = mean_f, color = author_type ))+ 
  geom_smooth(method = "lm",formula = y ~x, se = FALSE, fullrange=TRUE, size = 1.75) +
  geom_point(size = 3.5)+
  theme_void() +
  geom_label(data = label_num, aes(label=label), fill = NA,
             x = 2001, y = 51, hjust= 0, vjust = 0,  label.size = NA,
             family = "Oswald Bold", size = 7, color = "white" ,
             inherit.aes = FALSE) +
  geom_label(data = label_f, aes(label=label), fill = NA,
             y = 89, x = 2020, hjust= 1, vjust = 1,  label.size = NA,
             family = "Oswald Bold", size = 7, color = "#4ADABF" ,
             inherit.aes = FALSE) +
  geom_label(data = label_l, aes(label=label), fill = NA,
             y = 82, x = 2020, hjust= 1, vjust = 1,  label.size = NA,
             family = "Oswald Bold", size = 7, color = "white" ,
             inherit.aes = FALSE) +
  facet_wrap(toupper(organisation_name)~., nrow = 1)+
  scale_color_manual(values = col_pal) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "white") +
  scale_x_continuous(breaks = c(2000, 2020)) +
  scale_y_continuous(breaks = c(0, 100)) +
  labs(title = "PERCENTAGE OF FEMALE AUTHORS: FIRST VS. LAST (2000 - 2020)", 
       x = "YEAR", y = "PERCENTAGE") +
  theme(
    legend.position = "none",
    axis.text.x = element_text(family = "Oswald", size = 18, colour = "grey60"),
    axis.text.y = element_text(family = "Oswald", size = 18, colour = "grey60"),
    strip.text = element_text(family = "Oswald Bold", size = 18, colour = charcoal, margin = margin(b= 1)),
    axis.title.x = element_text(family = "Oswald Bold", size = 18, colour = "grey60"),
    axis.title.y = element_text(family = "Oswald Bold", size = 18, colour = "grey60", angle = 90),
    panel.spacing = unit(1.9, "lines"),
    panel.background = element_rect(fill = "#4E4BD8"),
    plot.margin = unit(c(0,0.5,0,0), "cm"),
    plot.title = element_text(face = "bold", family = "Oswald", hjust = 0, margin = margin(t = 40,b = 25), size = 24, color = charcoal),
    plot.subtitle = element_text(family = "Oswald", hjust = 0, margin = margin(b = 10, t = 30),colour = charcoal, size = 17),
    plot.caption =element_text(family = "Oswald ExtraLight", face = "italic", size = 6.5, hjust = 1, margin = margin(t = 5)))




# OVER TIME PROFESSORS & NON-PROFESSORS BY FACULTY ----------------------------------------------------------------

# data
prof_data_l <- prof_data %>% 
  pivot_longer(!organisation_name:position,names_to = "year", values_to = "perc") %>% 
  mutate(position = toupper(position),
         year = as.numeric(year),
         organisation_name = case_when(organisation_name == "Phil.-Hist. Fakultät" ~ "Humanities &\nSocial Sciences",                 
                                       organisation_name =="Phil.-Nat. Fakultät"  ~"Natural Sciences",                
                                       organisation_name == "Medizinische Fakultät"  ~ "Medicine",              
                                       organisation_name == "Juristische Fakultät" ~ "Law",                
                                       organisation_name == "Fakultät für Psychologie"  ~"Psychology",           
                                       organisation_name == "Theologische Fakultät"  ~ "Theology",                
                                       organisation_name == "Wirtschaftswissenschaftliche Fakultät" ~ "Business &\nEconomics"))



prof_data_l$organisation_name <- factor(prof_data_l$organisation_name, levels = sort(unique(prof_data_l$organisation_name)))

label_num <- data.frame(organisation_name= unique(prof_data_l$organisation_name), label= c(rep("", 6), "50%"))
label_p <- data.frame(organisation_name= unique(prof_data_l$organisation_name), label= c(rep("", 6), "ACADEMIC STAFF"))
label_m <- data.frame(organisation_name= unique(prof_data_l$organisation_name), label= c(rep("", 6), "PROFESSORSHIPS"))



# plot
charcoal <-  "#31424B"
col_pal <- c("#4ADABF", "white")

pA3 <- prof_data_l %>% 
  ggplot(aes(x = year, y = perc, color = position))+ 
  geom_smooth(method = "lm",formula = y ~x, se = FALSE, fullrange=TRUE, size = 1.75) +
  geom_point(size = 3.5)+
  theme_void() +
  geom_label(data = label_num, aes(label=label), fill = NA,
             x = 2001, y = 51, hjust= 0, vjust = 0,  label.size = NA,
             family = "Oswald Bold", size = 7, color = "white" ,
             inherit.aes = FALSE) +
  geom_label(data = label_p, aes(label=label), fill = NA,
             y = 89, x = 2020, hjust= 1, vjust = 1,  label.size = NA,
             family = "Oswald Bold", size = 7, color = "#4ADABF" ,
             inherit.aes = FALSE) +
  geom_label(data = label_m, aes(label=label), fill = NA,
             y = 82, x = 2020, hjust= 1, vjust = 1,  label.size = NA,
             family = "Oswald Bold", size = 7, color = "white" ,
             inherit.aes = FALSE) +
  facet_wrap(toupper(organisation_name)~., nrow = 1)+
  scale_color_manual(values = col_pal) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "white") +
  scale_x_continuous(breaks = c(2000, 2020), limits = c(2000,2020)) +
  scale_y_continuous(breaks = c(0, 100), limits = c(0,100)) +
  labs(title = "PERCENTAGE OF FEMALES EMPLOYED: ACADEMIC STAFF VS. PROFESSORSHIPS (2013 - 2019)", 
       x = "YEAR", y = "PERCENTAGE",
       caption = "Data from the University of Basel. Lines represent the predictions from linear regressions") +
  theme(
    legend.position = "none",
    axis.text.x = element_text(family = "Oswald", size = 18, colour = "grey60"),
    axis.text.y = element_text(family = "Oswald", size = 18, colour = "grey60"),
    strip.text = element_text(family = "Oswald Bold", size = 18, colour = charcoal, margin = margin(b= 1)),
    axis.title.x = element_text(family = "Oswald Bold", size = 18, colour = "grey60"),
    axis.title.y = element_text(family = "Oswald Bold", size = 18, colour = "grey60", angle = 90),
    panel.spacing = unit(1.9, "lines"),
    panel.background = element_rect(fill = "#4E4BD8"),
    plot.margin = unit(c(0,0.5,0,0), "cm"),
    plot.title = element_text(face = "bold", family = "Oswald", hjust = 0, margin = margin(t = 40,b = 25), size = 24, color = charcoal),
    plot.subtitle = element_text(family = "Oswald", hjust = 0, margin = margin(b = 10, t = 30),colour = charcoal, size = 17),
    plot.caption =element_text(family = "Oswald ExtraLight", size = 20, hjust = 1, margin = margin(t = 20, b = 3, r = 3)))

```

## The Publication Gender Gap at the University of Basel

```{r, echo=FALSE, eval=TRUE, fig.width = 20, fig.height= 10}

#pA <- pA1/pA2/pA3
#pA

fig_authors<- pA1

fig_authors

```



```{r, echo=FALSE, eval=TRUE}


prop_fem_total <- author_dat %>%
  ungroup() %>% 
  summarise(prop_female = mean(mean_f, na.rm = TRUE))

prop_fem <- author_dat %>%
  filter(year %in% c(2020, 2000)) %>%
  group_by(year) %>%
  summarise(prop_female = mean(mean_f, na.rm = TRUE))

prop_fem_inst <- author_dat %>%
  filter(year %in% c(2020, 2000)) %>%
  group_by(year, organisation_name) %>%
  summarise(prop_female = mean(mean_f, na.rm = TRUE)) %>% 
  group_by(organisation_name) %>% 
  mutate(diff = diff(prop_female)) 
  
prop_fem_inst_nyr <- author_dat %>%
  group_by(organisation_name) %>%
  summarise(prop_female = mean(mean_f, na.rm = TRUE))


```

- The figure above shows evidence of a publication gender gap at the University of Basel; across all faculties and time, there were overal fewer publications involving females (`r round(prop_fem_total$prop_female, 0)`%) relative to males (`r 100-round(prop_fem_total$prop_female, 0)`%); notwithstanding, the figure suggests a general upward trend in the percentage of publications involving females over the last two decades, while in 2000 the percentage of female authors was `r round(prop_fem$prop_female[1], 0)`%, in 2020 this percentage was `r round(prop_fem$prop_female[2], 0)`%.

- The overall levels as well as temporal trends vary, however, markedly between faculties: some faculties show a greater percentage of publications involving females across the full 20-year period (2000-2020), most notably the Faculty of Humanities and Social Sciences (`r round(prop_fem_inst_nyr$prop_female[prop_fem_inst_nyr$organisation_name == "Humanities &\nSocial Sciences"], 0)`%), the Faculty of Psychology (`r round(prop_fem_inst_nyr$prop_female[prop_fem_inst_nyr$organisation_name == "Psychology"], 0)`%), and the Faculty of Law (`r round(prop_fem_inst_nyr$prop_female[prop_fem_inst_nyr$organisation_name == "Law"], 0)`%). Concerning temporal trends, some faculties have seen a greater increase, most notably the Faculty of Psychology which saw an increase between 2000 to 2020 of `r round(prop_fem_inst$diff[prop_fem_inst$organisation_name == "Psychology" & prop_fem_inst$year == 2000], 0)`%. 

## Linking the Publication and Employment Gender Gaps: The Role of Seniority

We also investigated the publication gap as a function of authorship position by distinguishing between first and last authors (excluding single author publications). First authorship is typically (but not always) associated with junior academic roles in publications involving multiple authors, while last authorship typically cues seniority. Consequently, distinguishing between first and last authorship allows us to consider trends as a function of seniority or career stage. When considering these analysis and, particularly, comparing between faculties, it should be noted that determination of authorship order follows different practices across fields (e.g., alphabetical order in economics). Below, we show the breakdown of the publication gap by first vs. last author (upper panel) along with the percentage employment of females in professorships vs. academic staff (lower panel). 

```{r, echo=FALSE, eval=TRUE, fig.width = 20, fig.height= 19}

fig_authors_staff <- pA2/pA3
fig_authors_staff 

```

- As can be seen in the upper panel the figure above, there is a general trend for higher female proportions in first authors than last authors; this pattern matches the larger prevalence of females in junior (academic staff) vs. senior (professorship) roles that is evident in the lower panel. 

- As presented above for the overall percentage of publications, the temporal trends show overall increases over time, with patterns varying somewhat betwen faculties. Crucially, the temporal trends observed are not always parallel for first and last authors (cf. Natural Sciences, Psychology, Theology) suggesting that the dynamics underlying academic productivity can differ somewhat for the different career stages, at least for some faculties. 

### Key findings

- We identified a persistent and sizeable publication gender gap at the University of Basel between 2000-2020. 

- The publication gender gap has decreased over the last 20 years, albeit at different rates across the 7 faculties that compose the University of Basel. 

- The publication gender gap is larger and more persistant for last (i.e., senior authors) for some but not all faculties, suggesting that different dynamics are at play across faculties and career stages. 

## Limitations and Outlook

To be included after group discussion and input from Diversity Office (Oct 26th, 2021).

## References 

Odic, D., & Wojcik, E. H. (2020). The publication gender gap in psychology. *American Psychologist, 75*(1), 92–103. [http://doi.org/10.1037/amp0000480](http://doi.org/10.1037/amp0000480)