---
title: "Report"
description: Visualizing Diversity in Research Outputs at the University of Basel
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(ggtext)
library(patchwork)

pub_data <- read_csv("1_data/unibas_diversity.csv")

```

## Executive Summary

XXXX 


## Background

#### Why does diversity matter?

Achieving gender equality (SDG 5) and reducing inequalities (SDG10) are central parts of the [Sustainable Development Goals (SDGs)](https://sdgs.un.org/goals). Universities can play a key role in advancing these goals by empowering women and disadvantaged minorities to allow them to make contributions to scientific innovation and research. The University of Basel is strongly committed to monitoring and increasing diversity of its staff and student population in a number of efforts spearheaded by its [Diversity Office](https://www.unibas.ch/en/University/Administration-Services/Vice-President-s-Office-for-People-and-Culture/Diversity-and-Inclusion/Diversity.html). 

#### Why does monitoring of research outputs matter?

Monitoring diversity is a first step towards identifying inequalities and potential solutions to reduce them. As a consequence, there have been continued efforts on the part of the Diversity Office of the University of Basel to monitor diversity, for example, in the form of an [Equal opportunities monitoring](https://www.unibas.ch/en/University/Administration-Services/Vice-President-s-Office-for-People-and-Culture/Diversity-and-Inclusion/Diversity/Diversity-Management/Monitoring.html). However, the current monitoring focuses on the gender gap in employment and does not fully capture other aspects of equal opportunities. The monitoring of publication outputs can be helpful as an additional indicator that can better reflect access to resources (e.g., time, funding) or other inequalities that are otherwise not easily quantifiable and go beyond employment. 

Further, past monitoring has focused on gender but diversity should, ideally, be conceptualized through several dimensions. The tools of data science, such as the ones we use here, offer complementary approaches to capturing different dimensions of diversity, such as nationality or age that are not readily available for monitoring by the Diversity Office. 

### Goals of the Report
- visualize the diversity in research outputs at the University Basel
- provide ideas for future monitoring of diversity at the University of Basel, including future methodological approaches or data collection efforts
- discuss whether the monitoring of research outputs can help guide future interventions aimed at increasing diversity at the University of Basel

## Methods 

#### Research Output of the University of Basel

We use data from a database listing the University of Basel's research output. We focus on data between 2000 and 2020. RATIONALE FOR YEARS () comparability with previous work that used similar ranges (Odic & Wodjc, 2019), availability for earlier years is very limited.  We focus on faculties not associated institutes because there is larger number but vast majority has few publications that make it difficult to consider trends over time (link to organigram figure?)

There are a total 67'269 publications including several different categories: 43'997 journal articles, 11'959 book chapters, 3'454 books (edited and authored), 2'014 proceedings articles, 1'671 theses, and 4'174 other items including news articles and discussion papers.   

In a previous project, CDS members identified faculty-level organizational units within the University of Basel that we use to assign publications to specific organizational unites (i.g., faculties). 

#### Diversity Labeling from Author First Names

We use information concerning diversity derived from the authors' first names. 

We recruit the services [genderize.io](https://genderize.io/), [agify.io](https://agify.io/), [nationalize.io](https://nationalize.io/) to determine the authors' probable gender, age, and nation, respectively. Estimates are provided based on the services' databases that include hundreds of thousands of confirmed mappings on the internet between first names and these three characteristics. Additional details are provided on the [Data](data.html).   

We have processed the data in the following way to simplify our analysis. For gender and age, we only use the probable value, irrespective of the number of instances or the assigned probability. For nation, we determine whether any of the three most probable nations include a "western" nation defined as geographic Europe, the US, Canada, Australia, and New Zealand. 



The data is shared via this <a href="https://github.com/cdsbasel/Diversity_hackathon"><b>Github</b></a> repository. 

### Disclaimer
The following analysis uses first names to estimate the gender of the author. This classification system assigns names to either "female" or "male". While we would like to consider all genders, our analysis is therefore limited to these two categories.

### Gender diversity

```{r, echo=FALSE, eval=TRUE, fig.width = 15, fig.height= 12}
gender_inst_time <- pub_data %>% 
  select(year, gender_proportion, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:gender_proportion,names_to = "organisation_name", values_to = "applies") %>%
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Faculty of\nTheology",
                                       organisation_name == "Org-Faculty of Science" ~ "Faculty of\nScience",
                                       organisation_name == "Org-Faculty of Law" ~ "Faculty of\nLaw",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Faculty of\nHumanties & Soc. Sci.",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Faculty of\nMedicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Faculty of\nBusiness & Econ.",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Faculty of\nPsychology")) %>%
  filter(!is.na(organisation_name)) %>% # ignoring non-faculty organizations
  group_by(organisation_name, year) %>%
  summarise(FEMALE = 100*mean(gender_proportion, na.rm = TRUE),
            MALE = 100 - FEMALE) %>%
  pivot_longer(FEMALE:MALE, names_to = "gender", values_to = "publication")

gender_inst_time$gender = factor(gender_inst_time$gender,levels = c("MALE","FEMALE"))


# Plot

charcoal <-  "#31424B"

label_num <- data.frame(organisation_name= unique(gender_inst_time$organisation_name), label= c("50%", rep("", 6)))
label_f <- data.frame(organisation_name= unique(gender_inst_time$organisation_name), label= c("FEMALE", rep("", 6)))
label_m <- data.frame(organisation_name= unique(gender_inst_time$organisation_name), label= c("MALE", rep("", 6)))

pA1 <- ggplot(gender_inst_time, aes(x= year, y=publication, fill=gender)) +
  geom_bar(width = 1,position = "stack", stat = "identity") +
  scale_fill_manual(values = c("FEMALE" = "#4ADABF", "MALE" = "#4E4BD8"))+
  geom_hline(yintercept = 50, linetype = "dashed", color = "white") +
  theme_void() +
  geom_label(data = label_num, aes(label=label), fill = NA,
             x = 2001, y = 51, hjust= 0, vjust = 0,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  geom_label(data = label_f, aes(label=label), fill = NA,
             y = 5, x = 2010, hjust=0.5,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  geom_label(data = label_m, aes(label=label), fill = NA,
             y = 95, x = 2010, hjust=0.5,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  scale_x_continuous(breaks = c(2000, 2020)) +
  scale_y_continuous(expand = c(0.01,0), breaks = c(0, 100)) +
  facet_wrap(.~ toupper(organisation_name), nrow = 1) +
  labs(title = "PERCENTAGE OF FEMALE AUTHORS", subtitle = "OVER TIME BY FACULTY: ALL AUTHORSHIPS COMBINED", x = "YEAR", y = "PERC.") +
  theme(legend.position = "none",
        axis.text.x = element_text(family = "Arial", size = 9, colour = "grey60"),
        axis.text.y = element_text(family = "Arial", size = 9, colour = "grey60"),
        panel.spacing = unit(1, "lines"),
        strip.text = element_text(family = "Arial Bold", size = 11.5, colour = charcoal),
        axis.title.x = element_text(family = "Arial Bold", size = 10, colour = "grey60"),
        axis.title.y = element_text(family = "Arial Bold", size = 10, colour = "grey60", angle = 90),
        plot.title = element_text(face = "bold", family = "Arial", hjust = 0, margin = margin(t = 15,b = 25), size = 20, color = charcoal),
        plot.subtitle = element_text(family = "Arial", hjust = 0.008, margin = margin(b = 10), size = 17, color = charcoal),
        plot.caption =element_text(family = "Arial ExtraLight", face = "italic", size = 6.5, hjust = 1, margin = margin(t = 5)))



# OVER TIME FEMALE 1ST and LAST AUTHOR BY FACULTY ----------------------------------------------------------------

# data
author_first <- pub_data %>%
  filter(n_authors > 1) %>%
  select(year, gender_first, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:gender_first,names_to = "organisation_name", values_to = "applies") %>%
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Faculty of\nTheology",
                                       organisation_name == "Org-Faculty of Science" ~ "Faculty of\nScience",
                                       organisation_name == "Org-Faculty of Law" ~ "Faculty of\nLaw",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Faculty of\nHumanties & Soc. Sci.",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Faculty of\nMedicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Faculty of\nBusiness & Econ.",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Faculty of\nPsychology")) %>%
  filter(!is.na(organisation_name)) %>% # ignoring non-faculty organizations
  group_by(organisation_name, year) %>%
  summarise(mean_f = 100*mean(gender_first, na.rm = TRUE))  %>%
  mutate(author_type = "FIRST AUTHOR")



author_last <- pub_data %>%
  filter(n_authors > 1) %>%
  select(year, gender_last, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:gender_last,names_to = "organisation_name", values_to = "applies") %>%
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Faculty of\nTheology",
                                       organisation_name == "Org-Faculty of Science" ~ "Faculty of\nScience",
                                       organisation_name == "Org-Faculty of Law" ~ "Faculty of\nLaw",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Faculty of\nHumanties & Soc. Sci.",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Faculty of\nMedicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Faculty of\nBusiness & Econ.",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Faculty of\nPsychology")) %>%
  filter(!is.na(organisation_name)) %>% # ignoring non-faculty organizations
  group_by(organisation_name, year) %>%
  summarise(mean_f = 100*mean(gender_last, na.rm = TRUE))  %>%
  mutate(author_type = "LAST AUTHOR")

author_dat <- bind_rows(author_first, author_last)


label_num <- data.frame(organisation_name= unique(author_dat$organisation_name), label= c("50%", rep("", 6)))
label_f <- data.frame(organisation_name= unique(author_dat$organisation_name), label= c("FIRST AUTHOR", rep("", 6)))
label_m <- data.frame(organisation_name= unique(author_dat$organisation_name), label= c("LAST AUTHOR", rep("", 6)))




# plot

charcoal <-  "#31424B"
col_pal <- c("#4ADABF", "white")

pA2 <- author_dat %>%
  ggplot(aes(x = year, y = mean_f, color = author_type ))+
  geom_line(size = 1.2 ) +
  geom_point(size = 2)+
  theme_void() +
  geom_label(data = label_num, aes(label=label), fill = NA,
             x = 2001, y = 51, hjust= 0, vjust = 0,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  geom_label(data = label_f, aes(label=label), fill = NA,
             y = 95, x = 2001, hjust= 0, vjust = 1,  label.size = NA,
             family = "Arial Bold", size = 5, color = "#4ADABF" ,
             inherit.aes = FALSE) +
  geom_label(data = label_m, aes(label=label), fill = NA,
             y = 89, x = 2001, hjust= 0, vjust = 1,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  facet_wrap(toupper(organisation_name)~., nrow = 1)+
  scale_color_manual(values = col_pal) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "white") +
  scale_x_continuous(breaks = c(2000, 2020)) +
  scale_y_continuous(expand = c(0.01,0), breaks = c(0, 100)) +
  labs(title = "OVER TIME BY FACULTY: FIRST VERSUS LAST AUTHOR" , x = "YEAR", y = "PERC.",
       caption = "Data from the UNIBAS publication database.") +
  theme(legend.position = "none",
        axis.text.x = element_text(family = "Arial", size = 10, colour = "grey60"),
        axis.text.y = element_text(family = "Arial", size = 10, colour = "grey60",  margin = margin(l = 5, r = 5)),
        axis.title.x = element_text(family = "Arial", size = 10, colour = "grey60"),
        axis.title.y = element_text(family = "Arial", size = 10, colour = "grey60", angle = 90),
        panel.spacing = unit(2, "lines"),
        panel.background = element_rect(fill = "#4E4BD8"),
        plot.margin = unit(c(0,0.5,0,0), "cm"),
        strip.text.x = element_text(family = "Arial", size = 12, colour = charcoal, margin = margin(b = 3)),
        panel.border  = element_rect(colour = charcoal, fill = "NA", size = 0.75, linetype = "solid"),
        plot.title = element_text(family = "Arial", hjust = 0, margin = margin(b = 10, t = 30),colour = charcoal, size = 17),
        plot.caption =element_text(family = "Arial", face = "italic", size = 6.5, hjust = 1, margin = margin(t = 5)))



pA <- pA1/pA2

pA
```

```{r, echo=FALSE, eval=TRUE}



prop_fem <- author_dat %>%
  filter(year %in% c(2020, 2000)) %>%
  group_by(year) %>%
  summarise(prop_female = mean(mean_f, na.rm = TRUE))

prop_fem_inst <- author_dat %>%
  filter(year %in% c(2020, 2000)) %>%
  group_by(year, organisation_name) %>%
  summarise(prop_female = mean(mean_f, na.rm = TRUE)) %>% 
  group_by(organisation_name) %>% 
  mutate(diff = diff(prop_female)) 
  

prop_fem_inst_nyr <- author_dat %>%
  group_by(organisation_name) %>%
  summarise(prop_female = mean(mean_f, na.rm = TRUE))


```

<!-- Women are systematically underrepresented in academia. Whilst this inequality varies across academic departments, there has been an general upward trend in the proportion of female authors from the University of Basel over the last two decades, starting at `r round(prop_fem$prop_female[1], 2)` in 2000 and reaching `r round(prop_fem$prop_female[2], 2)` in 2020. This trend is also reflected in the percentage of first and last female authors. In this section, we offer a by-faculty insight into these developments over time. In particular, the prize for most improved goes to the Faculties of Humanities and Social Sciences, with Psychology and Medicine as runners up. These improvements are less noticeable in the Faculties of Science, Law, and Business & Economics, and the trend is reversed for the Faculty of Theology. However, this latter case may not be so informative as the sample of publications is considerably smaller. -->



- Considering the publications at the University of Basel over the last 20 years, in some faculties a greater percentage of authors are females in publications ranging from the Humanities and Social Sciences (`r round(prop_fem_inst_nyr$prop_female[prop_fem_inst_nyr$organisation_name == "Faculty of\nHumanties & Soc. Sci."], 2)`%) to Business and Economics  (`r round(prop_fem_inst_nyr$prop_female[prop_fem_inst_nyr$organisation_name == "Faculty of\nBusiness & Econ."], 2)`%).

- There has been an general upward trend in the proportion of female authors from the University of Basel over the last two decades who have published. While in 2000 `r round(prop_fem$prop_female[1], 2)`% of authors were female, this share has increased to `r round(prop_fem$prop_female[2], 2)`% in 2020.

- For some faculties this increase was more noticeable. For instance in the Faculty of Psychology the increase of percentage points between 2000 to 2020 was highest with `r round(prop_fem_inst$diff[prop_fem_inst$organisation_name == "Faculty of\nPsychology" & prop_fem_inst$year == 2000], 2)`%, while the Faculty of Law had the lowest increase between 2000 to 2020 with a difference of `r round(prop_fem_inst$diff[prop_fem_inst$organisation_name == "Faculty of\nLaw" & prop_fem_inst$year == 2000], 2)`%. Publication data for the Faculty of Theology allows for less confident statements but suggests a decrease in the proportion of female authors.

- In addition to the general author proportions, we investigated authorship positions by looking at first and last authors' gender. There is a general trend for higher female proportions in first authors than last authors. When comparing between faculties, it should be noted however, that authorship position assignment may follow different practices in different fields.

### Key points
- In the last 20 years, considerably more than half of the authors in the publications invesitgated irrespective of faculties, are male.

- This gender gap in publications has decreased for most faculties over the last 20 years. 

- Generally, in the publications looked at, proportion of women among first authors was higher than among last authors. 



## Limitations and Outlook


## Conclusion


## References 



