---
title: "Diversity at the University of Basel"
description: Diversity at the University of Basel (2000-2020)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(ggtext)
library(patchwork)
library(ggplot2)
pub_data <- read_csv("1_data/unibas_diversity.csv")



```

## Background

## Methods 

## Gender Diversity

```{r}


```


## Nationality

```{r}

# data
perc_all <- pub_data %>% 
  select(year, nation_proportion, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:nation_proportion,names_to = "organisation_name", values_to = "applies") %>% 
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Faculty of\nTheology",
                                       organisation_name == "Org-Faculty of Science" ~ "Faculty of\nScience",
                                       organisation_name == "Org-Faculty of Law" ~ "Faculty of\nLaw",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Faculty of\nHumanties & Soc. Sci.",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Faculty of\nMedicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Faculty of\nBusiness & Econ.",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Faculty of\nPsychology")) %>% 
  filter(!is.na(organisation_name))  %>% 
  summarize(mean_perc_west = mean(nation_proportion, na.rm=T))
perc_all <- perc_all*100
```
  
&nbsp; &nbsp;&nbsp;&nbsp;Across the seven faculties at the University of Basel, overall we find the majority of authors (`r round(perc_all, digits=2)`%) who published between 2000 and 2020 to be classified as being of western nationality. For some faculties (e.g., Humanities & Social Sciences, Law, and Theology) we find no substantial change of the nationality composition over time, while for other faculties (e.g., Medicine, Psychology, and Science) the percentage of non-western authors seems to have decreased over time.

&nbsp;&nbsp;&nbsp;&nbsp;Taking a more detailed look at author nationality by authorship position on the publication, overall we find no marked differences between first and last authors with regards to the composition of (non) western authors. Put differently, the nationality pattern over time is comparable for first and last authors, and this is the case for each of the seven faculties. The patterns we see for Faculty of Theology may be confounded by the small number of publications on which the classification of authors' nationality was based.


```{r, fig.width = 16, fig.height = 11 }

# data
nation_inst_time <- pub_data %>% 
  select(year, nation_proportion, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:nation_proportion,names_to = "organisation_name", values_to = "applies") %>% 
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Faculty of\nTheology",
                                       organisation_name == "Org-Faculty of Science" ~ "Faculty of\nScience",
                                       organisation_name == "Org-Faculty of Law" ~ "Faculty of\nLaw",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Faculty of\nHumanties & Soc. Sci.",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Faculty of\nMedicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Faculty of\nBusiness & Econ.",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Faculty of\nPsychology")) %>% 
  filter(!is.na(organisation_name)) %>% # ignoring non-faculty organizations
  group_by(organisation_name, year) %>% 
  summarise(WEST = 100*mean(nation_proportion, na.rm = TRUE),
            NONWEST = 100 - WEST) %>%
  pivot_longer(WEST:NONWEST, names_to = "nation", values_to = "publication")

nation_inst_time$nation = factor(nation_inst_time$nation,levels = c("WEST", "NONWEST"))

# Plot

charcoal <-  "#31424B"

label_num <- data.frame(organisation_name= unique(nation_inst_time$organisation_name), label= c("50%", rep("", 6)))
label_f <- data.frame(organisation_name= unique(nation_inst_time$organisation_name), label= c("NON-WESTERN", rep("", 6)))
label_m <- data.frame(organisation_name= unique(nation_inst_time$organisation_name), label= c("WESTERN", rep("", 6)))

pB1 <- ggplot(nation_inst_time, aes(x= year, y=publication, fill=nation)) +
  geom_bar(width = 1,position = "stack", stat = "identity") +
  scale_fill_manual(values = c( "WEST" = "#0f4c5c", "NONWEST" = "#6ede8a"))+
  geom_hline(yintercept = 50, linetype = "dashed", color = "white") +
  theme_void() +
  geom_label(data = label_num, aes(label=label), fill = NA,
             x = 2001, y = 51, hjust= 0, vjust = 0,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  geom_label(data = label_f, aes(label=label), fill = NA,
             y = 5, x = 2010, hjust=0.5,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  geom_label(data = label_m, aes(label=label), fill = NA,
             y = 95, x = 2010, hjust=0.5,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  scale_x_continuous(breaks = c(2000, 2020)) +
  scale_y_continuous(expand = c(0.01,0), breaks = c(0, 100)) +
  facet_wrap(.~ toupper(organisation_name), nrow = 1) +
  labs(title = "PERCENTAGE OF NON-WESTERN AUTHORS", subtitle = "OVER TIME BY FACULTY: ALL AUTHORSHIPS COMBINED", x = "YEAR", y = "PERC.") +
  theme(legend.position = "none",
        axis.text.x = element_text(family = "Arial", size = 9, colour = "grey60"),
        axis.text.y = element_text(family = "Arial", size = 9, colour = "grey60"),
        panel.spacing = unit(1, "lines"),
        strip.text = element_text(family = "Arial Bold", size = 11.5, colour = charcoal),
        axis.title.x = element_text(family = "Arial Bold", size = 10, colour = "grey60"),
        axis.title.y = element_text(family = "Arial Bold", size = 10, colour = "grey60", angle = 90),
        plot.title = element_text(face = "bold", family = "Arial", hjust = 0, margin = margin(t = 15,b = 25), size = 20, color = charcoal),
        plot.subtitle = element_text(family = "Arial", hjust = 0.008, margin = margin(b = 10), size = 17, color = charcoal),
        plot.caption =element_text(family = "Arial ExtraLight", face = "italic", size = 6.5, hjust = 1, margin = margin(t = 5)))



# OVER TIME FEMALE 1ST and LAST AUTHOR BY FACULTY ----------------------------------------------------------------

# data
author_first <- pub_data %>% 
  filter(n_authors > 1) %>% 
  select(year, nation_first, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:nation_first,names_to = "organisation_name", values_to = "applies") %>% 
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Faculty of\nTheology",
                                       organisation_name == "Org-Faculty of Science" ~ "Faculty of\nScience",
                                       organisation_name == "Org-Faculty of Law" ~ "Faculty of\nLaw",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Faculty of\nHumanties & Soc. Sci.",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Faculty of\nMedicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Faculty of\nBusiness & Econ.",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Faculty of\nPsychology")) %>% 
  filter(!is.na(organisation_name)) %>% # ignoring non-faculty organizations
  group_by(organisation_name, year) %>% 
  summarise(mean_f = 100-100*mean(nation_first, na.rm = TRUE))  %>% 
  mutate(author_type = "FIRST AUTHOR") 


author_last <- pub_data %>% 
  filter(n_authors > 1) %>% 
  select(year, nation_last, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:nation_last,names_to = "organisation_name", values_to = "applies") %>% 
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Faculty of\nTheology",
                                       organisation_name == "Org-Faculty of Science" ~ "Faculty of\nScience",
                                       organisation_name == "Org-Faculty of Law" ~ "Faculty of\nLaw",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Faculty of\nHumanties & Soc. Sci.",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Faculty of\nMedicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Faculty of\nBusiness & Econ.",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Faculty of\nPsychology")) %>% 
  filter(!is.na(organisation_name)) %>% # ignoring non-faculty organizations
  group_by(organisation_name, year) %>% 
  summarise(mean_f = 100-100*mean(nation_last, na.rm = TRUE))  %>% 
  mutate(author_type = "LAST AUTHOR") 

author_dat <- bind_rows(author_first, author_last) 


label_num <- data.frame(organisation_name= unique(author_dat$organisation_name), label= c("50%", rep("", 6)))
label_f <- data.frame(organisation_name= unique(author_dat$organisation_name), label= c("FIRST AUTHOR", rep("", 6)))
label_m <- data.frame(organisation_name= unique(author_dat$organisation_name), label= c("LAST AUTHOR", rep("", 6)))


# plot

charcoal <-  "#31424B"
col_pal <- c("#6ede8a", "white")

pB2 <- author_dat %>% 
  ggplot(aes(x = year, y = mean_f, color = author_type ))+ 
  geom_line(size = 1.2 ) +
  geom_point(size = 2)+
  theme_void() +
  geom_label(data = label_num, aes(label=label), fill = NA,
             x = 2001, y = 51, hjust= 0, vjust = 0,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  geom_label(data = label_f, aes(label=label), fill = NA,
             y = 95, x = 2001, hjust= 0, vjust = 1,  label.size = NA,
             family = "Arial Bold", size = 5, color = "#6ede8a" ,
             inherit.aes = FALSE) +
  geom_label(data = label_m, aes(label=label), fill = NA,
             y = 89, x = 2001, hjust= 0, vjust = 1,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  facet_wrap(toupper(organisation_name)~., nrow = 1)+
  scale_color_manual(values = col_pal) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "white") +
  scale_x_continuous(breaks = c(2000, 2020)) +
  scale_y_continuous(expand = c(0.01,0), breaks = seq(0, 100,100), limits = c(0,100)) +
  labs(title = "OVER TIME BY FACULTY: FIRST VERSUS LAST AUTHOR" , x = "YEAR", y = "PERC.",
       caption = "Data from the UNIBAS publication database.") +
  theme(legend.position = "none",
        axis.text.x = element_text(family = "Arial", size = 10, colour = "grey60"),
        axis.text.y = element_text(family = "Arial", size = 10, colour = "grey60",  margin = margin(l = 5, r = 5)),
        axis.title.x = element_text(family = "Arial Bold", size = 10, colour = "grey60"),
        axis.title.y = element_text(family = "Arial Bold", size = 10, colour = "grey60", angle = 90),
        panel.spacing = unit(2, "lines"),
        panel.background = element_rect(fill = "#0f4c5c"),
        plot.margin = unit(c(0,0.5,0,0), "cm"),
        strip.text.x = element_text(family = "Arial Bold", size = 12, colour = charcoal, margin = margin(b = 3)),
        panel.border  = element_rect(colour = charcoal, fill = "NA", size = 0.75, linetype = "solid"),
        plot.title = element_text(family = "Arial", hjust = 0, margin = margin(b = 10, t = 30),colour = charcoal, size = 17),
        plot.caption =element_text(family = "Arial Bold", face = "italic", size = 6.5, hjust = 1, margin = margin(t = 5)))

pB <- pB1/pB2
pB

```

## Limitations and Outlook


## Conclusion


## References 



