---
title: "Diversity at the University of Basel"
description: Diversity at the University of Basel (2000-2020)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(ggtext)
library(patchwork)

pub_data <- read_csv("1_data/unibas_diversity.csv")
```

## Background

## Methods

## Gender Diversity

```{r, echo=FALSE, eval=TRUE, fig.width = 15, fig.height= 12}
gender_inst_time <- pub_data %>% 
  select(year, gender_proportion, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:gender_proportion,names_to = "organisation_name", values_to = "applies") %>%
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Faculty of\nTheology",
                                       organisation_name == "Org-Faculty of Science" ~ "Faculty of\nScience",
                                       organisation_name == "Org-Faculty of Law" ~ "Faculty of\nLaw",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Faculty of\nHumanties & Soc. Sci.",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Faculty of\nMedicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Faculty of\nBusiness & Econ.",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Faculty of\nPsychology")) %>%
  filter(!is.na(organisation_name)) %>% # ignoring non-faculty organizations
  group_by(organisation_name, year) %>%
  summarise(FEMALE = 100*mean(gender_proportion, na.rm = TRUE),
            MALE = 100 - FEMALE) %>%
  pivot_longer(FEMALE:MALE, names_to = "gender", values_to = "publication")

gender_inst_time$gender = factor(gender_inst_time$gender,levels = c("MALE","FEMALE"))


# Plot

charcoal <-  "#31424B"

label_num <- data.frame(organisation_name= unique(gender_inst_time$organisation_name), label= c("50%", rep("", 6)))
label_f <- data.frame(organisation_name= unique(gender_inst_time$organisation_name), label= c("FEMALE", rep("", 6)))
label_m <- data.frame(organisation_name= unique(gender_inst_time$organisation_name), label= c("MALE", rep("", 6)))

pA1 <- ggplot(gender_inst_time, aes(x= year, y=publication, fill=gender)) +
  geom_bar(width = 1,position = "stack", stat = "identity") +
  scale_fill_manual(values = c("FEMALE" = "#4ADABF", "MALE" = "#4E4BD8"))+
  geom_hline(yintercept = 50, linetype = "dashed", color = "white") +
  theme_void() +
  geom_label(data = label_num, aes(label=label), fill = NA,
             x = 2001, y = 51, hjust= 0, vjust = 0,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  geom_label(data = label_f, aes(label=label), fill = NA,
             y = 5, x = 2010, hjust=0.5,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  geom_label(data = label_m, aes(label=label), fill = NA,
             y = 95, x = 2010, hjust=0.5,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  scale_x_continuous(breaks = c(2000, 2020)) +
  scale_y_continuous(expand = c(0.01,0), breaks = c(0, 100)) +
  facet_wrap(.~ toupper(organisation_name), nrow = 1) +
  labs(title = "PERCENTAGE OF FEMALE AUTHORS", subtitle = "OVER TIME BY FACULTY: ALL AUTHORSHIPS COMBINED", x = "YEAR", y = "PERC.") +
  theme(legend.position = "none",
        axis.text.x = element_text(family = "Arial", size = 9, colour = "grey60"),
        axis.text.y = element_text(family = "Arial", size = 9, colour = "grey60"),
        panel.spacing = unit(1, "lines"),
        strip.text = element_text(family = "Arial Bold", size = 11.5, colour = charcoal),
        axis.title.x = element_text(family = "Arial Bold", size = 10, colour = "grey60"),
        axis.title.y = element_text(family = "Arial Bold", size = 10, colour = "grey60", angle = 90),
        plot.title = element_text(face = "bold", family = "Arial", hjust = 0, margin = margin(t = 15,b = 25), size = 20, color = charcoal),
        plot.subtitle = element_text(family = "Arial", hjust = 0.008, margin = margin(b = 10), size = 17, color = charcoal),
        plot.caption =element_text(family = "Arial ExtraLight", face = "italic", size = 6.5, hjust = 1, margin = margin(t = 5)))



# OVER TIME FEMALE 1ST and LAST AUTHOR BY FACULTY ----------------------------------------------------------------

# data
author_first <- pub_data %>%
  filter(n_authors > 1) %>%
  select(year, gender_first, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:gender_first,names_to = "organisation_name", values_to = "applies") %>%
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Faculty of\nTheology",
                                       organisation_name == "Org-Faculty of Science" ~ "Faculty of\nScience",
                                       organisation_name == "Org-Faculty of Law" ~ "Faculty of\nLaw",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Faculty of\nHumanties & Soc. Sci.",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Faculty of\nMedicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Faculty of\nBusiness & Econ.",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Faculty of\nPsychology")) %>%
  filter(!is.na(organisation_name)) %>% # ignoring non-faculty organizations
  group_by(organisation_name, year) %>%
  summarise(mean_f = 100*mean(gender_first, na.rm = TRUE))  %>%
  mutate(author_type = "FIRST AUTHOR")



author_last <- pub_data %>%
  filter(n_authors > 1) %>%
  select(year, gender_last, `Org-Swiss Tropical and Public Health Institute`: `Org-Basel Institute on Governance`) %>%
  pivot_longer(!year:gender_last,names_to = "organisation_name", values_to = "applies") %>%
  filter(applies == 1) %>% #only select those rows that could be classified
  mutate(organisation_name = case_when(organisation_name == "Org-Faculty of Theology" ~ "Faculty of\nTheology",
                                       organisation_name == "Org-Faculty of Science" ~ "Faculty of\nScience",
                                       organisation_name == "Org-Faculty of Law" ~ "Faculty of\nLaw",
                                       organisation_name == "Org-Faculty of Humanties and Social Sciences" ~ "Faculty of\nHumanties & Soc. Sci.",
                                       organisation_name == "Org-Faculty of Medicine"  ~ "Faculty of\nMedicine",
                                       organisation_name == "Org-Faculty of Business and Economics" ~ "Faculty of\nBusiness & Econ.",
                                       organisation_name == "Org-Faculty of Psychology" ~ "Faculty of\nPsychology")) %>%
  filter(!is.na(organisation_name)) %>% # ignoring non-faculty organizations
  group_by(organisation_name, year) %>%
  summarise(mean_f = 100*mean(gender_last, na.rm = TRUE))  %>%
  mutate(author_type = "LAST AUTHOR")

author_dat <- bind_rows(author_first, author_last)


label_num <- data.frame(organisation_name= unique(author_dat$organisation_name), label= c("50%", rep("", 6)))
label_f <- data.frame(organisation_name= unique(author_dat$organisation_name), label= c("FIRST AUTHOR", rep("", 6)))
label_m <- data.frame(organisation_name= unique(author_dat$organisation_name), label= c("LAST AUTHOR", rep("", 6)))




# plot

charcoal <-  "#31424B"
col_pal <- c("#4ADABF", "white")

pA2 <- author_dat %>%
  ggplot(aes(x = year, y = mean_f, color = author_type ))+
  geom_line(size = 1.2 ) +
  geom_point(size = 2)+
  theme_void() +
  geom_label(data = label_num, aes(label=label), fill = NA,
             x = 2001, y = 51, hjust= 0, vjust = 0,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  geom_label(data = label_f, aes(label=label), fill = NA,
             y = 95, x = 2001, hjust= 0, vjust = 1,  label.size = NA,
             family = "Arial Bold", size = 5, color = "#4ADABF" ,
             inherit.aes = FALSE) +
  geom_label(data = label_m, aes(label=label), fill = NA,
             y = 89, x = 2001, hjust= 0, vjust = 1,  label.size = NA,
             family = "Arial Bold", size = 5, color = "white" ,
             inherit.aes = FALSE) +
  facet_wrap(toupper(organisation_name)~., nrow = 1)+
  scale_color_manual(values = col_pal) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "white") +
  scale_x_continuous(breaks = c(2000, 2020)) +
  scale_y_continuous(expand = c(0.01,0), breaks = c(0, 100)) +
  labs(title = "OVER TIME BY FACULTY: FIRST VERSUS LAST AUTHOR" , x = "YEAR", y = "PERC.",
       caption = "Data from the UNIBAS publication database.") +
  theme(legend.position = "none",
        axis.text.x = element_text(family = "Arial", size = 10, colour = "grey60"),
        axis.text.y = element_text(family = "Arial", size = 10, colour = "grey60",  margin = margin(l = 5, r = 5)),
        axis.title.x = element_text(family = "Arial", size = 10, colour = "grey60"),
        axis.title.y = element_text(family = "Arial", size = 10, colour = "grey60", angle = 90),
        panel.spacing = unit(2, "lines"),
        panel.background = element_rect(fill = "#4E4BD8"),
        plot.margin = unit(c(0,0.5,0,0), "cm"),
        strip.text.x = element_text(family = "Arial", size = 12, colour = charcoal, margin = margin(b = 3)),
        panel.border  = element_rect(colour = charcoal, fill = "NA", size = 0.75, linetype = "solid"),
        plot.title = element_text(family = "Arial", hjust = 0, margin = margin(b = 10, t = 30),colour = charcoal, size = 17),
        plot.caption =element_text(family = "Arial", face = "italic", size = 6.5, hjust = 1, margin = margin(t = 5)))



pA <- pA1/pA2

pA
````

```{r, echo=FALSE, eval=TRUE}



prop_fem <- author_dat %>%
  filter(year %in% c(2020, 2000)) %>%
  group_by(year) %>%
  summarise(prop_female = mean(mean_f, na.rm = TRUE))

prop_fem_inst <- author_dat %>%
  filter(year %in% c(2020, 2000)) %>%
  group_by(year, organisation_name) %>%
  summarise(prop_female = mean(mean_f, na.rm = TRUE)) %>% 
  group_by(organisation_name) %>% 
  mutate(diff = diff(prop_female)) 
  

prop_fem_inst_nyr <- author_dat %>%
  group_by(organisation_name) %>%
  summarise(prop_female = mean(mean_f, na.rm = TRUE))


```

<!-- Women are systematically underrepresented in academia. Whilst this inequality varies across academic departments, there has been an general upward trend in the proportion of female authors from the University of Basel over the last two decades, starting at `r round(prop_fem$prop_female[1], 2)` in 2000 and reaching `r round(prop_fem$prop_female[2], 2)` in 2020. This trend is also reflected in the percentage of first and last female authors. In this section, we offer a by-faculty insight into these developments over time. In particular, the prize for most improved goes to the Faculties of Humanities and Social Sciences, with Psychology and Medicine as runners up. These improvements are less noticeable in the Faculties of Science, Law, and Business & Economics, and the trend is reversed for the Faculty of Theology. However, this latter case may not be so informative as the sample of publications is considerably smaller. -->

### Disclaimer
The following analysis uses first names to estimate the gender of the author. This classification system assigns names to either "female" or "male". While we would like to consider all genders, our analysis is therefore limited to these two categories.

### Gender diversity

- Considering the publications at the University of Basel over the last 20 years, in some faculties a greater percentage of authors are females in publications ranging from the Humanities and Social Sciences (`r round(prop_fem_inst_nyr$prop_female[prop_fem_inst_nyr$organisation_name == "Faculty of\nHumanties & Soc. Sci."], 2)`%) to Business and Economics  (`r round(prop_fem_inst_nyr$prop_female[prop_fem_inst_nyr$organisation_name == "Faculty of\nBusiness & Econ."], 2)`%).

- There has been an general upward trend in the proportion of female authors from the University of Basel over the last two decades who have published. While in 2000 `r round(prop_fem$prop_female[1], 2)`% of authors were female, this share has increased to `r round(prop_fem$prop_female[2], 2)`% in 2020.

- For some faculties this increase was more noticeable. For instance in the Faculty of Psychology the increase of percentage points between 2000 to 2020 was highest with `r round(prop_fem_inst$diff[prop_fem_inst$organisation_name == "Faculty of\nPsychology" & prop_fem_inst$year == 2000], 2)`%, while the Faculty of Law had the lowest increase between 2000 to 2020 with a difference of `r round(prop_fem_inst$diff[prop_fem_inst$organisation_name == "Faculty of\nLaw" & prop_fem_inst$year == 2000], 2)`%. Publication data for the Faculty of Theology allows for less confident statements but suggests a decrease in the proportion of female authors.

- In addition to the general author proportions, we investigated authorship positions by looking at first and last authors' gender. There is a general trend for higher female proportions in first authors than last authors. When comparing between faculties, it should be noted however, that authorship position assignment may follow different practices in different fields.

### Key points
- In the last 20 years, considerably more than half of the authors in the publications invesitgated irrespective of faculties, are male.

- This gender gap in publications has decreased for most faculties over the last 20 years. 

- Generally, in the publications looked at, proportion of women among first authors was higher than among last authors. 


## Nationality

```{r}

```

## Limitations and Outlook


## Conclusion


## References
